---
title: "PA1_Template.Rmd"
author: "Bob Gardner"
date: "March 25, 2016"
output:
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Course Project 1:  Reproducible Research

### Loading and Preprocessing the Data
Read in files and transform as will be needed downstream.

```{r loadtransform}
activity<-read.csv("activity.csv")
activity$date<-as.Date(activity$date)
```

### What is mean total number of steps taken per day?

Calculate the total number of steps for each day, plot in histogram, and mean/median.

```{r daysteps}
dailySums<-tapply(activity$steps,activity$date,sum)
hist(dailySums,breaks=11, main = "Total Number of Steps Each Day", xlab="Number of Steps in a Day")
avg<-as.integer(mean(dailySums,na.rm=TRUE))
med<-median(dailySums,na.rm=TRUE)
```

The mean of the total number of steps taken per day is `r avg` steps.

The median of the total number of steps taken per day is `r med` steps.

### What is the average daily activity pattern?

Prepare and plot time series of 5-minute intervals averaged across days and find max average steps interval.

```{r intervalavgs}
dailyAvg<-tapply(activity$steps,activity$interval,mean,na.rm=TRUE)
plot(row.names(dailyAvg),dailyAvg,type="l", ylab = "number of steps per 5 minute interval", xlab = "5 minute interval during the 24 hour day")
max<-names(which.max(dailyAvg))
```

The 5-minute interval, on average across all the days in the dataset, which contains the maximum number of steps is `r max`.

### Imputing missing values

Calculate the number of missing values in the dataset and prepare new dataset with NAs replaced with interval means.

```{r impute}
NACount<-sum(is.na(activity$steps))
imputeNA<-numeric()
dailyAvg<-aggregate(steps ~ interval, data = activity, FUN = mean)
for (i in 1:nrow(activity)){
	obs<-activity[i,]
	if (is.na(obs$steps)){
		steps<-subset(dailyAvg, interval == obs$interval)$steps
	} else {
		steps<-obs$steps
	}
	imputeNA<-c(imputeNA,steps)
}
activity2.0<-activity
activity2.0$steps<-imputeNA
dailySums2.0<-tapply(activity2.0$steps,activity2.0$date,sum)
hist(dailySums2.0,breaks=11, main = "Total Number of Steps Each Day, Imputed", xlab="Number of Steps in a Day")
avg2.0<-as.integer(mean(dailySums2.0))
med2.0<-as.integer(median(dailySums2.0))
```

The calculated total number of missing values in the dataset is `r NACount`.

The mean using imputed means to replace NAs of the total number of steps taken per day is `r avg2.0` steps.

The median using imputed means to replace NAs of the total number of steps taken per day is `r med2.0` steps.

**There is no difference between the imputed estimates from the first part of the assignment.**

### Are there differences in activity patterns between weekdays and weekends?

Create new factor variable in imputed dataset by weekend and weekday and panel plot.

```{r}
days<-weekdays(activity$date, abbreviate = FALSE)
dayLevel<-vector()
for (i in 1:nrow(activity2.0)){
	if (days[i] == "Saturday") {
		dayLevel[i]<-"weekend"
	} else if (days[i] == "Sunday") {
		dayLevel[i]<-"weekend"
	} else {
		dayLevel[i]<-"weekday"
	}
}
activity2.0$dayLevel<-dayLevel
activity2.0$dayLevel<-factor(activity2.0$dayLevel)
dayTypeSteps<-aggregate(steps ~ interval + dayLevel, data = activity2.0, mean)
require(lattice)
xyplot(steps ~ interval|dayLevel,dayTypeSteps, type = "l", layout = c(1,2), xlab = "interval", ylab = "Number of Steps by Interval, Average")
```

When comparing the weekend to the weekday averages, the subject is less active until later in the morning, more active throughout the day, and remains active later into the evening.





